---
title: "03 · Sample selection with weasel"
---

The analytic sample keeps every respondent observed on at least three of the seven analysis waves, and estimates under full-information maximum likelihood. This vignette expresses that rule as an explicit, named, auditable scenario with [`weasel`](https://github.com/siardv/weasel), shows the audit outputs the article draws on, and demonstrates the machinery live on synthetic data.

## Semantics first

`weasel` counts a respondent as observed at a wave when a row with that `(id, wave)` pair exists in the long data. The study's notion of observed is substantive: at least one focal measure (BMI, fruit and vegetable consumption, or sport hours) reported at that wave. The bridge is a presence filter applied before planning:

```{r}
#| eval: false
presence <- liss %>%
  dplyr::filter(
    t %in% analysis_waves,
    !is.na(bmi) | !is.na(fv) | !is.na(pa)
  )
```

Rows with a wave record but no focal measurement therefore do not count as observed, exactly as in the pipeline's own rule.

## The rule as a scenario

Three defaults ship with `weasel_plan()` (`anchored_strict`, `anchored_balanced`, `lenient_info_max`); the study's rule is a fourth, custom one. Within the full seven-wave span, at least three observed waves means at most four missing; endpoints are not required, and interior-gap limits are set non-binding.

```{r}
#| eval: false
ws_plan <- weasel::weasel_plan(
  presence[c("id", "t", "female", "age")],
  id = "id", wave = "t", span = "full",
  scenarios = data.frame(
    scenario = "min3_of7",
    require_endpoints = FALSE,
    max_missing = 4L,
    n_gap_max = 5L,
    max_gap_max = 5L
  )
)
analysis_ids <- unique(weasel::weasel_apply(ws_plan, "min3_of7")$id)
```

The pipeline (chunk `cb112` of `analysis/reciprocal-health-dynamics.R`) runs this plan alongside its own one-liner and asserts `setequal()` between the two id sets, so the transcript proves at every render that the scenario and the rule are the same selection: 5,676 respondents.

## The audits the article uses

Three diagnostics turn the rule from a stipulation into an argued choice.

**Tolerance sensitivity.** `weasel_sensitivity()` sweeps the minimum-wave tolerance and reports the retained sample at each level, which is the sample-size trade-off behind choosing three as the threshold:

```{r}
#| eval: false
weasel::weasel_sensitivity(
  ws_plan,
  require_endpoints = FALSE, max_missing = 0:6,
  n_gap_max = 5L, max_gap_max = 5L
)
```

**Attrition selectivity.** `weasel_selectivity()` compares retained and excluded respondents on covariates via standardised mean differences, the empirical check on whether completeness-based selection skews the sample that the missing-at-random discussion in the article leans on:

```{r}
#| eval: false
weasel::weasel_selectivity(ws_plan, "min3_of7")
```

**Prose justification.** `weasel_justify_subset()` generates methods-section text stating the window, the constraints, the retained count, and the coverage achieved, regenerable from the same inputs:

```{r}
#| eval: false
cat(weasel::weasel_justify_subset(ws_plan, "min3_of7"))
```

`scripts/02_select_sample.R` runs all of the above against the real panel and writes the id list and the justification to `output/`.

## Live demonstration on synthetic data

Everything above executes here on generated data with the study's exact scenario, so the outputs can be inspected without any LISS access.

```{r}
#| eval: !expr requireNamespace("weasel", quietly = TRUE)
set.seed(42)
d <- do.call(rbind, lapply(1:400, function(i) {
  w <- sort(sample(1:7, sample(1:7, 1)))
  data.frame(id = i, t = w, female = stats::rbinom(1, 1, 0.5),
             age = stats::rnorm(1, 50, 15))
}))

sc <- data.frame(
  scenario = "min3_of7", require_endpoints = FALSE,
  max_missing = 4L, n_gap_max = 5L, max_gap_max = 5L
)
p <- weasel::weasel_plan(d, id = "id", wave = "t", span = "full", scenarios = sc)
p
```

```{r}
#| eval: !expr requireNamespace("weasel", quietly = TRUE)
weasel::weasel_print_table(
  weasel::weasel_sensitivity(p, require_endpoints = FALSE, max_missing = 0:6,
                             n_gap_max = 5L, max_gap_max = 5L),
  title = "sample size by minimum-wave tolerance"
)
weasel::weasel_print_table(
  weasel::weasel_selectivity(p, "min3_of7"),
  title = "retained vs excluded"
)
cat(weasel::weasel_justify_subset(p, "min3_of7"))
```

```{r}
#| eval: !expr requireNamespace("weasel", quietly = TRUE)
sub <- weasel::weasel_apply(p, "min3_of7")
identical(
  sort(unique(sub$id)),
  sort(as.integer(names(which(table(d$id) >= 3))))
)
```

The final `identical()` is the same equivalence the pipeline asserts on the real data: the scenario is the rule.

## Interactive exploration

Before committing to any rule, the scope pipeline maps the participation landscape:

```{r}
#| eval: false
weasel::set_weasel_scope(presence, "id", "t", gap = 1)
weasel::evaluate_weasel_scope()
weasel::weasel_summarize_waves()
weasel::weasel_print_table(weasel::weasel_filter_wave_summary(), n = 10)
weasel::weasel_scope_info()
weasel::weasel_clear_scope()
```

Panels on non-consecutive schedules (biennial waves, waves recorded as years) are handled with `grid = "observed"` in both pipelines.
