---
title: "S6 · Results and reproduction"
---

This vignette renders every released table from `tables/` and `output/`, and closes with the exact reproduction ladder. It requires no LISS data.

```{r}
tbl <- function(f) utils::read.csv(file.path("..", "tables", f), check.names = FALSE)
out <- function(f) utils::read.csv(file.path("..", "output", f), check.names = FALSE)
k <- function(x, ...) knitr::kable(x, digits = 3, ...)
```

## Model comparison: CLPM versus RI-CLPM

```{r}
k(tbl("clpm_vs_riclpm_fit.csv"), caption = "Pooled sample")
k(tbl("clpm_vs_riclpm_fit_grouped.csv"), caption = "Three-group (multigroup) models")
```

The RI-CLPM dominates on every descriptive index and by thousands of AIC and BIC points in both the pooled and the multigroup fits: the covariation among the constructs is predominantly between persons.

## Between-person (trait-level) associations

```{r}
bp <- tbl("between_person_covariances.csv")
k(bp, caption = "Random-intercept variances and covariances (unstandardised)")

v <- setNames(bp$est[bp$lhs == bp$rhs], bp$lhs[bp$lhs == bp$rhs])
cv <- bp[bp$lhs != bp$rhs, ]
cv$r <- cv$est / sqrt(v[cv$lhs] * v[cv$rhs])
k(cv[, c("lhs", "rhs", "est", "r", "z", "pvalue")],
  caption = "Trait covariances with implied correlations")
```

Trait BMI relates negatively to trait activity, is essentially unrelated to the fruit-and-vegetable trait, and trait activity relates positively to the diet trait.

## Within-person dynamics (pooled, time-invariant)

```{r}
k(tbl("riclpm_dynamics_coefficients_pooled.csv"),
  caption = "Autoregressive and cross-lagged coefficients")
```

Carry-over is moderate and weakest for diet; among the cross-lags, the BMI-PA pair is reciprocal, with both paths negative and small.

## Socioeconomic moderation

```{r}
k(tbl("focal_covariance_by_ses.csv"),
  caption = "Standardised trait BMI-PA covariance by stratum")
k(tbl("focal_covariance_boot.csv"),
  caption = "Delta-method and bootstrap percentile intervals (R = 1,000)")
k(tbl("moderation_lrt.csv"), caption = "Free versus equal across strata (scaled LRT)")
```

The focal covariance is negative and significant in every stratum, tightly clustered, and a model holding it equal is preferred: no socioeconomic moderation.

## Construction audit and robustness

The audit of the construction-stage BMI processing (S4) and the sensitivity refit that excludes every affected observation (S5):

```{r}
k(out("bmi_construction_audit_summary.csv"),
  caption = "Construction audit: per-wave person-wave counts by category")
k(out("bmi_construction_audit_magnitudes.csv"),
  caption = "Size bands of the altered values (kg/m²)")
k(out("bmi_construction_audit_rawonly.csv"),
  caption = "Raw-only person-waves: absent from the merged panel, by wave")
k(tbl("bmi_sensitivity_between.csv"),
  caption = "Sensitivity: between-person (co)variances, flagged BMI set missing")
k(tbl("bmi_sensitivity_focal_by_ses.csv"),
  caption = "Sensitivity: standardised trait BMI-PA covariance by stratum")
k(tbl("bmi_sensitivity_moderation_lrt.csv"),
  caption = "Sensitivity: moderation test (scaled LRT)")
```

The sensitivity estimates reproduce the primary analysis throughout: the article's conclusions do not depend on the corrected or filled values.

## Measurement evidence

```{r}
k(out("medication_mokken.csv"),
  caption = "Mokken scale analysis of the medication count, per wave and pooled")
k(out("frve_reliability_raw.csv"),
  caption = "FV items: per-wave correlation and Spearman-Brown (raw six-point items)")
k(out("frve_reliability_pooled.csv"), caption = "FV items: pooled reliability")
k(out("baseline_correlations.csv"), caption = "Baseline (wave 1) correlations")
k(out("adjacent_wave_stability.csv"),
  caption = "Adjacent-wave stabilities (lag-1 Pearson r): mean and range")
```

```{r}
cat(readLines(file.path("..", "output", "educ_other_count.txt")), sep = "\n")
cat(readLines(file.path("..", "output", "frve_mapping_deterministic.txt")), sep = "\n")
```

## Full parameter tables

Every parameter of the four fitted models, with standard errors, z statistics, p values, confidence intervals, and standardised values, ships as CSV; the pooled RI-CLPM is previewed here and the others are summarised by size.

```{r}
fp <- c(
  "full_parameters_riclpm_pooled.csv", "full_parameters_riclpm_grouped.csv",
  "full_parameters_riclpm_covariates.csv", "full_parameters_clpm_pooled.csv"
)
for (f in fp) cat(sprintf("- %s: %d parameter rows\n", f, nrow(tbl(f))))
k(head(tbl("full_parameters_riclpm_pooled.csv"), 12),
  caption = "Preview: pooled RI-CLPM parameters (first 12 of the full CSV)")
```

## Supporting tests and per-group detail

```{r}
k(tbl("time_invariance_lrt.csv"), caption = "Lagged paths free versus time-invariant")
k(tbl("riclpm_dynamics_coefficients.csv"),
  caption = "Per-group time-invariant dynamics")
k(tbl("ses_group_estimates.csv"), caption = "Per-group estimate overview")
k(tbl("riclpm_covariate_effects.csv"),
  caption = "Covariate effects on the random intercepts (standardised, fixed.x = TRUE)")
```

## Reproduction ladder

1. `remotes::install_deps()` from the compendium root (pulls `lissr` and `weasel` from GitHub per `DESCRIPTION`).
2. Obtain LISS access, then `make acquire` and `make panel` (S1 and S2), or place an existing `liss_merged_long.sav` in `data/` and verify it: `(cd data && shasum -a 256 -c CHECKSUMS.txt)`.
3. `make sample` for the standalone weasel audit (it also writes the id list the reliability script needs), then `make frve`, `make audit`, and `make mokken` for the measurement and construction evidence.
4. `make analysis` renders `analysis/run_all.md`, the chunk-by-chunk transcript ending in `sessionInfo()`; `make release-tables` syncs the released `tables/`. The first render takes roughly two hours, dominated by the bootstrap; the seeded, fingerprint-guarded cache makes subsequent renders take minutes.
5. `sh scripts/publish_site.sh` re-renders this site with the current transcript and publishes it.

Two assertions run inside the pipeline on every render: the `lissr` equivalisation must equal the in-line formula, and the `weasel` scenario must select exactly the pipeline's 5,676 respondents. A failed assertion stops the render rather than silently diverging, and the render driver additionally fails hard if the transcript completed without rewriting its result tables.

```{r}
sessionInfo()
```
