---
title: "RI-CLPM analysis transcript"
always_allow_html: true
output:
  github_document:
    toc: true
    html_preview: false
---

```{r setup, include=FALSE}
# echo every step's code, print messages and warnings, keep going on errors so
# the transcript stays complete, and send figures where github renders them.
knitr::opts_chunk$set(
  echo = TRUE, message = TRUE, warning = TRUE, error = TRUE,
  comment = "#>", fig.path = "run_all_files/figure-gfm/"
)
library(magrittr) # %>%, %<>%, %$%
# source the compendium helpers robustly: knitr evaluates this document with
# the working directory set to analysis/, so R/ sits one level up. fail loudly
# rather than degrading to whatever helper definitions the calling session
# happens to carry.
helper_dir <- if (dir.exists("R")) "R" else file.path("..", "R")
helper_files <- list.files(helper_dir, pattern = "[.][Rr]$", full.names = TRUE)
stopifnot(length(helper_files) > 0)
invisible(lapply(helper_files, source))

# pull the validated chunks out of reciprocal-health-dynamics.R by their
# "# ---- cb<id> ----" headers, in the order they should run.
src <- readLines("reciprocal-health-dynamics.R", warn = FALSE)
hdr <- grep("^# ---- cb[0-9]+ ----", src)
ids <- as.integer(sub("^# ---- cb([0-9]+) ----.*", "\\1", src[hdr]))
bounds <- c(hdr, length(src) + 1L)
chunk <- function(id) {
  k <- match(id, ids)
  src[(hdr[k] + 1L):(bounds[k + 1L] - 1L)]
}
run_ids <- c(
  51:57, 107, 111, 112:115, 117, 119, 122, 124,
  125, 135, 137, 146, 148, 157, 167, 169, 178, 180, 189,
  194:197, 206:208, 213, 224:226
)
```

## 1 data

```{r data}
data_dir <- liss_data_dir()

# seed the long frame from the merged backbone.
liss <- read_liss("liss_merged_long.sav", data_dir = data_dir)

# data provenance: file times of the merged backbone and the four source
# extracts it was assembled from. on archived copies these record the copy,
# not the original construction; the authoritative pin for the frozen inputs
# is data/CHECKSUMS.txt.
sav_files <- c(
  "liss_merged_long.sav", "liss_health.sav", "liss_sport.sav",
  "liss_income.sav", "liss_background.sav"
)
fs::file_info(fs::path(data_dir, sav_files)) %>%
  dplyr::transmute(
    file        = sav_files,
    time_source = ifelse(is.na(birth_time), "modified", "birth"),
    created     = dplyr::coalesce(birth_time, modification_time),
    age_days    = round(as.numeric(difftime(Sys.time(), created, units = "days")))
  ) %>%
  dplyr::arrange(created)
```

## 2 analysis steps

```{r steps, results='asis', echo=FALSE}
# emit one child chunk per step so knitr echoes its code and prints every
# intermediate result, not just the last value. objects persist across steps
# because the children share the global knit environment.
step_desc <- trimws(sub("^# ---- cb[0-9]+ ----[ ]?", "", src[hdr]))
names(step_desc) <- ids
for (k in seq_along(run_ids)) {
  id <- run_ids[k]
  title <- step_desc[as.character(id)]
  if (is.na(title) || title == "") title <- sprintf("step cb%d", id)
  cat(
    knitr::knit_child(
      text  = c(sprintf("### 2.%d %s", k, title), "", sprintf("```{r cb%d}", id), chunk(id), "```", ""),
      envir = knitr::knit_global(),
      quiet = TRUE
    ),
    sep = "\n"
  )
}
```

## 3 result tables

```{r tables}
stopifnot(all(c(paste0("BMI", 1:7), paste0("PA", 1:7), paste0("FV", 1:7)) %in% names(wide)))
cat(sprintf(
  "N = %d | SES groups: %s\n", nrow(wide),
  paste(names(table(wide$ses)), table(wide$ses), sep = "=", collapse = " / ")
))
dir.create("tables", showWarnings = FALSE)
# missing-data accounting under FIML: the effective N actually used by the
# likelihood, the realised coverage of the analysis variables, and the listwise
# N it replaces.
analysis_vars <- c(paste0("BMI", 1:7), paste0("PA", 1:7), paste0("FV", 1:7))
listwise_n <- sum(complete.cases(wide[analysis_vars]))

ri_fit <- all_model_fits$pooled$ri_clpm_pooled_fit
effective_n <- lavaan::lavInspect(ri_fit, "ntotal")
coverage <- diag(lavaan::lavInspect(ri_fit, "coverage"))

cat(sprintf(
  paste0(
    "FIML effective N = %d | listwise N = %d | panel N = %d\n",
    "realised coverage: %.1f%% to %.1f%% across the %d indicators (mean %.1f%%)\n"
  ),
  effective_n, listwise_n, nrow(wide),
  100 * min(coverage), 100 * max(coverage), length(coverage), 100 * mean(coverage)
))

# preliminary: CLPM vs RI-CLPM on an identical lag structure.
# the CLPM is nested in the RI-CLPM as the boundary case with the three
# random-intercept variances (and with them their covariances) fixed to zero;
# because that constraint sits on the boundary of the parameter space, the
# scaled chi-square difference test does not follow its nominal distribution
# (Stoel et al., 2006). compare on robust descriptive fit indices and on AIC
# and BIC instead.
fit_msrs <- c("cfi.robust", "tli.robust", "rmsea.robust", "srmr", "aic", "bic")

fit_compare <- function(clpm, riclpm) {
  m <- as.data.frame(rbind(
    CLPM      = lavaan::fitMeasures(clpm, fit_msrs),
    `RI-CLPM` = lavaan::fitMeasures(riclpm, fit_msrs)
  ))
  names(m) <- c("CFI", "TLI", "RMSEA", "SRMR", "AIC", "BIC")
  m <- rbind(m, difference = m["RI-CLPM", ] - m["CLPM", ])
  cbind(model = rownames(m), m)
}

# pooled comparison (headline preliminary)
clpm_vs_riclpm <- fit_compare(
  all_model_fits$pooled$clpm_pooled_fit,
  all_model_fits$pooled$ri_clpm_pooled_fit
)
utils::write.csv(clpm_vs_riclpm, "tables/clpm_vs_riclpm_fit.csv", row.names = FALSE)
knitr::kable(
  clpm_vs_riclpm,
  row.names = FALSE, digits = c(0, 3, 3, 3, 3, 0, 0),
  caption = "preliminary: CLPM vs RI-CLPM, pooled (robust fit, AIC, BIC)"
)

# grouped comparison (confirmation across SES groups)
clpm_vs_riclpm_grouped <- fit_compare(
  all_model_fits$grouped$clpm_grouped_fit,
  all_model_fits$grouped$ri_clpm_grouped_fit
)
utils::write.csv(
  clpm_vs_riclpm_grouped, "tables/clpm_vs_riclpm_fit_grouped.csv",
  row.names = FALSE
)
knitr::kable(
  clpm_vs_riclpm_grouped,
  row.names = FALSE, digits = c(0, 3, 3, 3, 3, 0, 0),
  caption = "preliminary: CLPM vs RI-CLPM, by SES group (robust fit, AIC, BIC)"
)

# pooled between-person (random-intercept) variances and covariances
ppe <- lavaan::parameterEstimates(all_model_fits$pooled$ri_clpm_pooled_fit)
between <- ppe[
  ppe$op == "~~" & ppe$lhs %in% c("riBMI", "riPA", "riFV"),
  c("lhs", "rhs", "est", "se", "z", "pvalue")
]
utils::write.csv(between, "tables/between_person_covariances.csv", row.names = FALSE)
knitr::kable(between, row.names = FALSE, caption = "between-person (co)variances")

# full parameter tables: every estimate with SE, z, p, 95% CI, and standardised
# values, for the pooled and grouped RI-CLPM, the covariate model, and the
# pooled CLPM benchmark (the tables the article's supplementary-material
# statement promises)
dump_params <- function(fit, file) {
  pe <- lavaan::parameterEstimates(fit, standardized = TRUE)
  utils::write.csv(pe, file.path("tables", file), row.names = FALSE)
  invisible(pe)
}
dump_params(all_model_fits$pooled$ri_clpm_pooled_fit, "full_parameters_riclpm_pooled.csv")
dump_params(all_model_fits$grouped$ri_clpm_grouped_fit, "full_parameters_riclpm_grouped.csv")
dump_params(all_model_fits$grouped$ri_clpm_grouped_covar_fit, "full_parameters_riclpm_covariates.csv")
dump_params(all_model_fits$pooled$clpm_pooled_fit, "full_parameters_clpm_pooled.csv")
cat("full parameter tables written to tables/full_parameters_*.csv\n")

# path estimates compared across SES groups
ses_estimates <- compare_ses_groups(
  all_model_fits$grouped$ri_clpm_grouped_fit,
  include_p_values = FALSE, hide_non_significant = FALSE, only_lagged_effect = FALSE
)
utils::write.csv(ses_estimates, "tables/ses_group_estimates.csv", row.names = FALSE)
knitr::kable(ses_estimates, caption = "path estimates by SES group")

# descriptive statistics for the wide panel
writeLines(
  as.character(describe(wide, exclude = "id", caption = "Descriptive statistics")),
  "tables/descriptives.html"
)

# model fit indices for the pooled RI-CLPM (uses the optional 'report' package)
if (requireNamespace("report", quietly = TRUE)) {
  fit_tables <- model_fit_table(all_model_fits$pooled$ri_clpm_pooled_fit)
  writeLines(as.character(fit_tables[[1]]), "tables/ri_clpm_fit_parameters.html")
  writeLines(as.character(fit_tables[[2]]), "tables/ri_clpm_fit_indices.md")
}

list.files("tables")
```

## 4 measurement invariance (interpretation)

Because the RI-CLPM fixes every factor loading to one, metric (equal-loading)
invariance holds by construction, which is enough to compare the structural
parameters across SES groups: the autoregressive and cross-lagged regressions
and the between-person random-intercept (co)variances. The across-SES comparison
in the analysis steps provides the formal structural test, whether the
regression paths can be held equal across groups; scalar (equal-intercept) and
strict (equal-residual) invariance are not required here and are reported only
for completeness.

## 5 moderation test

```{r moderation}
# formal test of the socioeconomic moderation: is the focal between-person
# covariance (the BMI and PA random intercepts) equal across SES groups? fit the
# grouped RI-CLPM with that one covariance held equal across groups, then compare
# it to the free grouped fit with a scaled likelihood-ratio test. the constrained
# model is nested in the free one (it fixes a single covariance equal across
# groups), so a scaled LRT is the right tool, unlike the non-nested CLPM vs
# RI-CLPM comparison.
free_fit <- all_model_fits$grouped$ri_clpm_grouped_fit

# constrain riBMI ~~ riPA equal across the three groups via one shared label;
# everything else stays free per group.
ri_clpm_syntax_eq <- sub(
  "riBMI ~~ riPA", "riBMI ~~ c(rc, rc, rc)*riPA",
  ri_clpm_syntax, fixed = TRUE
)
eq_fit <- lavaan::lavaan(
  model = ri_clpm_syntax_eq,
  data = wide,
  group = "ses",
  group.label = c("ses1", "ses2", "ses3"),
  int.ov.free = TRUE,
  meanstructure = TRUE,
  estimator = "MLR",
  se = "robust",
  missing = "fiml",
  fixed.x = FALSE
)

mod_lrt <- lavaan::lavTestLRT(
  free_fit, eq_fit,
  method = "satorra.2000",
  model.names = c("riBMI~~riPA free", "riBMI~~riPA equal")
)
lrt_df <- cbind(model = rownames(mod_lrt), as.data.frame(mod_lrt))
utils::write.csv(lrt_df, "tables/moderation_lrt.csv", row.names = FALSE)
knitr::kable(
  lrt_df, row.names = FALSE, digits = 3,
  caption = "moderation test: riBMI ~~ riPA free vs equal across SES (scaled LRT)"
)
```

```{r moderation-estimates}
# standardised focal covariance with 95% CIs, by SES group.
std <- lavaan::standardizedSolution(free_fit, level = 0.95)
focal <- std[
  std$op == "~~" &
    ((std$lhs == "riBMI" & std$rhs == "riPA") |
      (std$lhs == "riPA" & std$rhs == "riBMI")),
  c("group", "est.std", "se", "ci.lower", "ci.upper", "pvalue")
]
focal$SES <- factor(focal$group, levels = 1:3, labels = c("ses1", "ses2", "ses3"))
utils::write.csv(focal, "tables/focal_covariance_by_ses.csv", row.names = FALSE)
knitr::kable(
  focal[c("SES", "est.std", "se", "ci.lower", "ci.upper", "pvalue")],
  row.names = FALSE, digits = 3,
  caption = "standardised BMI-PA random-intercept covariance by SES group (95% CI)"
)
```

```{r moderation-bootstrap}
# bootstrap percentile CIs for the focal standardised covariance, per SES group.
# reuse tables/focal_covariance_boot.csv when a stored fit fingerprint still
# matches the current free_fit, skipping the slow refit; a changed fit re-runs the
# bootstrap automatically. delete the csv to force a fresh run regardless. boot_R
# is the draw count when it does run.
boot_R <- 1000L
boot_seed <- 20240501L
boot_csv <- "tables/focal_covariance_boot.csv"

# cache validity fingerprint. the cached intervals belong to one specific fit, so
# fingerprint free_fit by its focal point estimates (the bootstrapped quantity),
# its effective N and free-parameter count, and the bootstrap settings. reuse the
# cache on an exact match; a mismatch re-runs the bootstrap. a cache present with
# no fingerprint (the first render after this patch) is adopted and fingerprinted.
fp_file <- "tables/focal_covariance_boot.fingerprint"
fp_focal <- focal$est.std
names(fp_focal) <- as.character(focal$SES)
fp_focal <- fp_focal[order(names(fp_focal))]
boot_fingerprint <- paste(
  paste0("R=", boot_R),
  paste0("seed=", boot_seed),
  paste0("ntotal=", lavaan::lavInspect(free_fit, "ntotal")),
  paste0("npar=", lavaan::lavInspect(free_fit, "npar")),
  paste0("focal=", paste(names(fp_focal), signif(fp_focal, 6), sep = ":", collapse = ",")),
  sep = "|"
)
cached_fp <- if (file.exists(fp_file)) trimws(readLines(fp_file, warn = FALSE))[1] else NA_character_
reuse_cache <- file.exists(boot_csv) &&
  (is.na(cached_fp) || identical(cached_fp, boot_fingerprint))

if (reuse_cache) {
  if (is.na(cached_fp)) {
    cat("Adopting the existing bootstrap cache; recording the fit fingerprint.\n")
  } else {
    cat("Reusing cached bootstrap CIs (fit fingerprint matched).\n")
  }
  prev <- utils::read.csv(boot_csv, stringsAsFactors = FALSE)
  focal$boot.lower <- prev$boot.lower[match(as.character(focal$SES), prev$SES)]
  focal$boot.upper <- prev$boot.upper[match(as.character(focal$SES), prev$SES)]
  focal$n_boot <- prev$n_boot[1]
} else {
  cat("Running the bootstrap (no cache present, or the fit fingerprint changed).\n")
  focal_std_fun <- function(fit) {
    ss <- lavaan::standardizedSolution(fit)
    sel <- ss$op == "~~" &
      ((ss$lhs == "riBMI" & ss$rhs == "riPA") |
        (ss$lhs == "riPA" & ss$rhs == "riBMI"))
    out <- ss$est.std[sel]
    names(out) <- paste0("ses", ss$group[sel])
    out
  }
  ncpus <- max(1L, parallel::detectCores() - 1L)
  use_par <- if (.Platform$OS.type == "windows") "no" else "multicore"
  set.seed(boot_seed)
  boot_args <- list(object = free_fit, R = boot_R, FUN = focal_std_fun, parallel = use_par)
  if (use_par != "no") boot_args <- c(boot_args, list(ncpus = ncpus, iseed = boot_seed))
  boot_focal <- do.call(lavaan::bootstrapLavaan, boot_args)
  if (is.null(colnames(boot_focal))) colnames(boot_focal) <- c("ses1", "ses2", "ses3")
  boot_focal <- boot_focal[stats::complete.cases(boot_focal), , drop = FALSE]
  boot_ci <- t(apply(boot_focal, 2, stats::quantile, probs = c(0.025, 0.975), na.rm = TRUE))
  colnames(boot_ci) <- c("boot.lower", "boot.upper")
  focal$boot.lower <- boot_ci[as.character(focal$SES), "boot.lower"]
  focal$boot.upper <- boot_ci[as.character(focal$SES), "boot.upper"]
  focal$n_boot <- nrow(boot_focal)
}

focal_out <- focal[c("SES", "est.std", "ci.lower", "ci.upper", "boot.lower", "boot.upper", "n_boot")]
utils::write.csv(focal_out, boot_csv, row.names = FALSE)
# persist the fingerprint describing the fit these intervals belong to
writeLines(boot_fingerprint, fp_file)
knitr::kable(
  focal_out, row.names = FALSE, digits = 3,
  caption = paste(
    "standardised BMI-PA random-intercept covariance by SES group:",
    "delta-method vs bootstrap percentile 95% CIs"
  )
)
```

```{r moderation-coef-plot}
# coefficient plot (figure 2): point estimates with bootstrap percentile CIs,
# the more trustworthy interval for a standardised covariance.
ggplot2::ggplot(focal, ggplot2::aes(x = SES, y = est.std)) +
  ggplot2::geom_hline(yintercept = 0, linetype = "dashed") +
  ggplot2::geom_errorbar(
    ggplot2::aes(ymin = boot.lower, ymax = boot.upper), width = 0.15
  ) +
  ggplot2::geom_point(size = 3) +
  ggplot2::labs(
    x = "SES group",
    y = "standardised BMI-PA random-intercept covariance",
    title = "Between-person BMI-PA association by SES (bootstrap 95% CI)"
  ) +
  ggplot2::theme_bw()
```

## 6 RI-CLPM coefficients

The autoregressive and cross-lagged coefficients are reported from a time-invariant grouped RI-CLPM (lagged paths constrained equal across waves within each SES group, free across groups), so each effect is a single row per SES group; these are the invariant raw coefficients, so they are unstandardised. Dynamics are from the no-covariate model behind the moderation analysis. Covariate effects on the random intercepts come from a refit conditioning on the covariates (`fixed.x = TRUE`), standardised when that model's implied covariance is positive definite. The core dynamics models elsewhere keep `fixed.x = FALSE`; if a refit fails to converge, the relevant block falls back to the earlier free-per-wave / unstandardised form. A fully pooled time-invariant model (lagged paths equal across waves and across SES groups, justified by the across-SES structural invariance result) summarises the dynamics in nine unstandardised coefficients, reported alongside the per-group tables. A scaled likelihood-ratio test compares the free-per-wave grouped model with its time-invariant counterpart to check the time-invariance assumption. Estimates, intervals, and the test are written to `tables/`.

```{r coefficients}
# refinement 1: dynamics from a time-invariant grouped ri-clpm (lagged paths
# share labels across waves within each SES group, free across groups via
# group-specific label vectors), one row per effect, reported unstandardised (the
# constrained raw coefficients are the invariant quantities). refinement 2:
# covariate effects standardised from the fixed.x = TRUE refit when its implied
# covariance is positive definite. both fall back if a refit fails.
is_within <- function(v) grepl("^w(BMI|PA|FV)[1-7]$", v)
fam <- function(v) sub("^w(BMI|PA|FV)[1-7]$", "\\1", v)
star <- function(p) {
  ifelse(p <= 0.001, "***",
    ifelse(p <= 0.01, "**", ifelse(p <= 0.05, "*", ifelse(p <= 0.1, ".", ""))))
}
fmt_p <- function(p) ifelse(p < 0.001, "<0.001", formatC(p, format = "f", digits = 3))

# build a time-invariant variant of the no-covariate syntax by labelling the
# lagged paths and sharing those labels across waves
lagged_free <- paste(
  "wBMI2 + wPA2 + wFV2 ~ wBMI1 + wPA1 + wFV1",
  "wBMI3 + wPA3 + wFV3 ~ wBMI2 + wPA2 + wFV2",
  "wBMI4 + wPA4 + wFV4 ~ wBMI3 + wPA3 + wFV3",
  "wBMI5 + wPA5 + wFV5 ~ wBMI4 + wPA4 + wFV4",
  "wBMI6 + wPA6 + wFV6 ~ wBMI5 + wPA5 + wFV5",
  "wBMI7 + wPA7 + wFV7 ~ wBMI6 + wPA6 + wFV6",
  sep = "\n"
)
lagged_ti <- paste(vapply(1:6, function(t) {
  sprintf(paste(
    "wBMI%d ~ c(bb1,bb2,bb3)*wBMI%d + c(pb1,pb2,pb3)*wPA%d + c(fb1,fb2,fb3)*wFV%d",
    "wPA%d ~ c(bp1,bp2,bp3)*wBMI%d + c(pp1,pp2,pp3)*wPA%d + c(fp1,fp2,fp3)*wFV%d",
    "wFV%d ~ c(bf1,bf2,bf3)*wBMI%d + c(pf1,pf2,pf3)*wPA%d + c(ff1,ff2,ff3)*wFV%d",
    sep = "\n"
  ), t + 1L, t, t, t, t + 1L, t, t, t, t + 1L, t, t, t)
}, character(1)), collapse = "\n")
ri_clpm_syntax_ti <- sub(lagged_free, lagged_ti, ri_clpm_syntax, fixed = TRUE)

free_fit_ti <- NULL
if (grepl("c(bb1,bb2,bb3)*wBMI", ri_clpm_syntax_ti, fixed = TRUE)) {
  free_fit_ti <- tryCatch(
    lavaan::lavaan(
      model = ri_clpm_syntax_ti, fixed.x = FALSE, data = wide, group = "ses",
      group.label = c("ses1", "ses2", "ses3"), int.ov.free = TRUE,
      meanstructure = TRUE, estimator = "MLR", se = "robust", missing = "fiml"
    ),
    error = function(e) NULL
  )
}

lab_map <- c(
  bb = "BMI -> BMI", pp = "PA -> PA", ff = "FV -> FV",
  pb = "PA -> BMI", fb = "FV -> BMI", bp = "BMI -> PA",
  fp = "FV -> PA", bf = "BMI -> FV", pf = "PA -> FV"
)

if (!is.null(free_fit_ti) && isTRUE(lavaan::lavInspect(free_fit_ti, "converged"))) {
  dyn <- lavaan::parameterEstimates(free_fit_ti)
  dyn <- dyn[dyn$op == "~" & grepl("^(bb|pp|ff|pb|fb|bp|fp|bf|pf)[123]$", dyn$label), ]
  dyn <- dyn[!duplicated(dyn[c("group", "label")]), ]
  dyn$prefix <- sub("[123]$", "", dyn$label)
  dyn$group <- paste0("ses", dyn$group)
  dyn$path <- lab_map[dyn$prefix]
  dyn$block <- ifelse(dyn$prefix %in% c("bb", "pp", "ff"), "autoregressive", "cross-lagged")
  est_col <- "est"
  dyn_cap <- "time-invariant, unstandardised"
} else {
  dyn <- lavaan::standardizedSolution(free_fit)
  dyn <- dyn[dyn$op == "~" & is_within(dyn$lhs) & is_within(dyn$rhs), ]
  dyn$group <- paste0("ses", dyn$group)
  dyn$path <- paste(dyn$rhs, "->", dyn$lhs)
  dyn$block <- ifelse(fam(dyn$lhs) == fam(dyn$rhs), "autoregressive", "cross-lagged")
  est_col <- "est.std"
  dyn_cap <- "free per wave, standardised (time-invariant refit did not converge)"
}
dyn$sig <- star(dyn$pvalue)
dyn_tab <- dyn[, c("block", "group", "path", est_col, "se", "ci.lower", "ci.upper", "pvalue", "sig")]
dyn_tab <- dyn_tab[order(dyn_tab$block, dyn_tab$path, dyn_tab$group), ]
utils::write.csv(dyn_tab, "tables/riclpm_dynamics_coefficients.csv", row.names = FALSE)
dyn_tab$pvalue <- fmt_p(dyn_tab$pvalue)
keep <- setdiff(names(dyn_tab), "block")

knitr::kable(
  dyn_tab[dyn_tab$block == "autoregressive", keep],
  format = "markdown", row.names = FALSE, digits = 3,
  caption = paste0("autoregressive coefficients by SES group (", dyn_cap, ")")
)
knitr::kable(
  dyn_tab[dyn_tab$block == "cross-lagged", keep],
  format = "markdown", row.names = FALSE, digits = 3,
  caption = paste0("cross-lagged coefficients by SES group (", dyn_cap, ")")
)

# refinement 3: fully pooled time-invariant dynamics. lagged paths held equal
# across waves and across SES groups (justified by the across-SES structural
# invariance result), collapsing each effect to one unstandardised coefficient.
# built like the per-group time-invariant syntax but with scalar labels, which a
# grouped lavaan model constrains across groups as well; same config as free_fit.
lagged_ti_pooled <- paste(vapply(1:6, function(t) {
  sprintf(paste(
    "wBMI%d ~ bb*wBMI%d + pb*wPA%d + fb*wFV%d",
    "wPA%d ~ bp*wBMI%d + pp*wPA%d + fp*wFV%d",
    "wFV%d ~ bf*wBMI%d + pf*wPA%d + ff*wFV%d",
    sep = "\n"
  ), t + 1L, t, t, t, t + 1L, t, t, t, t + 1L, t, t, t)
}, character(1)), collapse = "\n")
ri_clpm_syntax_ti_pooled <- sub(lagged_free, lagged_ti_pooled, ri_clpm_syntax, fixed = TRUE)

free_fit_ti_pooled <- NULL
if (grepl("bb*wBMI", ri_clpm_syntax_ti_pooled, fixed = TRUE)) {
  # scalar labels across groups intentionally impose the cross-group equality
  # that pools the coefficients; lavaan's advisory warning about single labels
  # in a multigroup model is expected here and suppressed as such
  free_fit_ti_pooled <- tryCatch(
    suppressWarnings(lavaan::lavaan(
      model = ri_clpm_syntax_ti_pooled, fixed.x = FALSE, data = wide, group = "ses",
      group.label = c("ses1", "ses2", "ses3"), int.ov.free = TRUE,
      meanstructure = TRUE, estimator = "MLR", se = "robust", missing = "fiml"
    )),
    error = function(e) NULL
  )
}

if (!is.null(free_fit_ti_pooled) && isTRUE(lavaan::lavInspect(free_fit_ti_pooled, "converged"))) {
  pool <- lavaan::parameterEstimates(free_fit_ti_pooled)
  pool <- pool[pool$op == "~" & pool$label %in% names(lab_map), ]
  pool <- pool[!duplicated(pool$label), ]
  pool$path <- lab_map[pool$label]
  pool$block <- ifelse(pool$label %in% c("bb", "pp", "ff"), "autoregressive", "cross-lagged")
  pool$sig <- star(pool$pvalue)
  pool_tab <- pool[, c("block", "path", "est", "se", "ci.lower", "ci.upper", "pvalue", "sig")]
  pool_tab <- pool_tab[order(pool_tab$block, pool_tab$path), ]
  utils::write.csv(pool_tab, "tables/riclpm_dynamics_coefficients_pooled.csv", row.names = FALSE)
  pool_tab$pvalue <- fmt_p(pool_tab$pvalue)
  knitr::kable(
    pool_tab, format = "markdown", row.names = FALSE, digits = 3,
    caption = "pooled coefficients (time-invariant across waves and SES groups, unstandardised)"
  )
} else {
  cat("\nThe pooled time-invariant refit did not converge; pooled coefficients are not shown.\n")
}

# refinement 4: scaled likelihood-ratio test of the time-invariance assumption.
# free_fit (the free-per-wave grouped RI-CLPM from the moderation section, cb167)
# and free_fit_ti (the per-group time-invariant model above) share estimator MLR,
# se robust, missing fiml, fixed.x = FALSE and int.ov.free, so free_fit_ti is
# nested in free_fit and the scaled diff test is valid. guard against future drift
# in those settings: if they diverge, refit a matching free-per-wave model.
ti_opt_keys <- c("estimator", "se", "missing", "fixed.x")
opts_match <- function(a, b) {
  isTRUE(all.equal(
    lavaan::lavInspect(a, "options")[ti_opt_keys],
    lavaan::lavInspect(b, "options")[ti_opt_keys]
  ))
}
ti_lrt_df <- NULL
ti_msg <- NULL
if (exists("free_fit") && !is.null(free_fit_ti) &&
  isTRUE(lavaan::lavInspect(free_fit_ti, "converged"))) {
  free_fit_cmp <- free_fit
  if (!opts_match(free_fit, free_fit_ti)) {
    free_fit_cmp <- tryCatch(
      lavaan::lavaan(
        model = ri_clpm_syntax, fixed.x = FALSE, data = wide, group = "ses",
        group.label = c("ses1", "ses2", "ses3"), int.ov.free = TRUE,
        meanstructure = TRUE, estimator = "MLR", se = "robust", missing = "fiml"
      ),
      error = function(e) NULL
    )
  }
  if (!is.null(free_fit_cmp) && isTRUE(lavaan::lavInspect(free_fit_cmp, "converged"))) {
    ti_lrt <- tryCatch(
      lavaan::lavTestLRT(
        free_fit_cmp, free_fit_ti, method = "satorra.2000",
        model.names = c("lagged paths free per wave", "lagged paths time-invariant")
      ),
      error = function(e) NULL
    )
    if (!is.null(ti_lrt)) {
      ti_lrt_df <- cbind(model = rownames(ti_lrt), as.data.frame(ti_lrt))
      utils::write.csv(ti_lrt_df, "tables/time_invariance_lrt.csv", row.names = FALSE)
      cn <- names(ti_lrt)
      ci <- grep("Chisq diff", cn, fixed = TRUE)
      di <- grep("Df diff", cn, fixed = TRUE)
      pp <- grep("Pr(>", cn, fixed = TRUE)
      if (length(ci) && length(di) && length(pp)) {
        rws <- which(!is.na(ti_lrt[[cn[ci[1]]]]))
        if (length(rws)) {
          r <- rws[length(rws)]
          ti_chisq <- ti_lrt[[cn[ci[1]]]][r]
          ti_dfd <- as.integer(round(ti_lrt[[cn[di[1]]]][r]))
          ti_pval <- ti_lrt[[cn[pp[1]]]][r]
          ti_verdict <- if (is.na(ti_pval)) {
            "the scaled difference test was not computable"
          } else if (ti_pval > 0.05) {
            "the data are consistent with holding the lagged paths time-invariant"
          } else {
            "the data argue against full time-invariance of the lagged paths"
          }
          ti_msg <- sprintf(
            "time-invariance LRT: scaled chi-square difference = %.2f, df = %d, p = %s; %s.",
            ti_chisq, ti_dfd, fmt_p(ti_pval), ti_verdict
          )
        }
      }
    }
  }
}

if (!is.null(ti_lrt_df)) {
  knitr::kable(
    ti_lrt_df, row.names = FALSE, digits = 3,
    caption = "time-invariance LRT: lagged paths free per wave vs time-invariant (scaled)"
  )
} else {
  cat("\nThe time-invariance LRT could not be computed; a required fit was unavailable.\n")
}

if (!is.null(ti_msg)) cat("\n", ti_msg, "\n", sep = "")

# covariate effects on the random intercepts, conditioning on the covariates
covar_fit <- tryCatch(
  lavaan::lavaan(
    model = ri_clpm_covar_syntax, fixed.x = TRUE, data = wide, group = "ses",
    group.label = c("ses1", "ses2", "ses3"), int.ov.free = TRUE,
    meanstructure = TRUE, estimator = "MLR", se = "robust", missing = "fiml"
  ),
  error = function(e) NULL
)
if (!is.null(covar_fit) && isTRUE(lavaan::lavInspect(covar_fit, "converged"))) {
  cv_std <- tryCatch(lavaan::standardizedSolution(covar_fit), error = function(e) NULL)
  if (!is.null(cv_std)) {
    cv <- cv_std
    cv_col <- "est.std"
    cv_cap <- "standardised"
  } else {
    cv <- lavaan::parameterEstimates(covar_fit)
    cv_col <- "est"
    cv_cap <- "unstandardised, covariate model not positive definite"
  }
  cv <- cv[cv$op == "~" & grepl("^ri(BMI|PA|FV)$", cv$lhs) & !is_within(cv$rhs), ]
  cv$group <- paste0("ses", cv$group)
  cv$path <- paste(cv$rhs, "->", cv$lhs)
  cv$sig <- star(cv$pvalue)
  cv_tab <- cv[, c("group", "path", cv_col, "se", "ci.lower", "ci.upper", "pvalue", "sig")]
  cv_tab <- cv_tab[order(cv_tab$path, cv_tab$group), ]
  utils::write.csv(cv_tab, "tables/riclpm_covariate_effects.csv", row.names = FALSE)
  cv_tab$pvalue <- fmt_p(cv_tab$pvalue)
  knitr::kable(
    cv_tab, format = "markdown", row.names = FALSE, digits = 3,
    caption = paste0("covariate effects on the random intercepts by SES group (", cv_cap, ", fixed.x = TRUE)")
  )
} else {
  cat("\nThe covariate-adjusted refit did not converge; covariate effects are not shown.\n")
}
```

## 7 bmi construction-audit sensitivity

The frozen input's BMI underwent construction-stage processing (unit-error
repairs, person-level height smoothing, and rare filling of missing inputs);
`scripts/05_bmi_construction_audit.R` quantifies this against the raw health
modules and writes person-wave flags. Here the flagged observations (categories
`value_changed` and `merged_only`) are set missing and the focal models are
refit; agreement with the primary estimates shows the conclusions do not ride
on the corrected values. The chunk skips gracefully when the flags file is
absent, since creating it requires the raw LISS modules.

```{r bmi-sensitivity}
flags_file <- fs::path(liss_data_dir(), "bmi_construction_flags.csv")
if (!file.exists(flags_file)) {
  cat("bmi_construction_flags.csv not found; run scripts/05_bmi_construction_audit.R first. skipping.\n")
} else {
  fl <- utils::read.csv(flags_file)
  fl <- fl[fl$category %in% c("value_changed", "merged_only"), ]
  wide_sens <- wide
  n_blanked <- 0L
  for (w in 1:7) {
    ids_w <- fl$nomem_encr[fl$wavenr == w]
    hit <- wide_sens$id %in% ids_w
    n_blanked <- n_blanked + sum(hit & !is.na(wide_sens[[paste0("BMI", w)]]))
    wide_sens[hit, paste0("BMI", w)] <- NA_real_
  }
  cat(sprintf(
    "blanked %d flagged BMI person-waves (of %d flag rows) in the analysis frame\n",
    n_blanked, nrow(fl)
  ))

  sens_pooled <- lavaan::lavaan(
    model = ri_clpm_syntax, fixed.x = FALSE, data = wide_sens,
    int.ov.free = TRUE, meanstructure = TRUE,
    estimator = "MLR", se = "robust", missing = "fiml"
  )
  pe_s <- lavaan::parameterEstimates(sens_pooled)
  bp_s <- pe_s[pe_s$op == "~~" & pe_s$lhs %in% c("riBMI", "riPA", "riFV") &
    pe_s$rhs %in% c("riBMI", "riPA", "riFV"),
  c("lhs", "rhs", "est", "se", "z", "pvalue")]
  utils::write.csv(bp_s, "tables/bmi_sensitivity_between.csv", row.names = FALSE)
  print(knitr::kable(bp_s,
    row.names = FALSE, digits = 3,
    caption = "sensitivity: between-person (co)variances with flagged BMI set missing"
  ))

  sens_grouped <- lavaan::lavaan(
    model = ri_clpm_syntax, fixed.x = FALSE, data = wide_sens, group = "ses",
    group.label = c("ses1", "ses2", "ses3"), int.ov.free = TRUE,
    meanstructure = TRUE, estimator = "MLR", se = "robust", missing = "fiml"
  )
  std_s <- lavaan::standardizedSolution(sens_grouped, level = 0.95)
  foc_s <- std_s[std_s$op == "~~" &
    ((std_s$lhs == "riBMI" & std_s$rhs == "riPA") |
      (std_s$lhs == "riPA" & std_s$rhs == "riBMI")),
  c("group", "est.std", "se", "ci.lower", "ci.upper", "pvalue")]
  foc_s$SES <- paste0("ses", foc_s$group)
  utils::write.csv(foc_s, "tables/bmi_sensitivity_focal_by_ses.csv", row.names = FALSE)
  print(knitr::kable(foc_s[c("SES", "est.std", "se", "ci.lower", "ci.upper", "pvalue")],
    row.names = FALSE, digits = 3,
    caption = "sensitivity: standardised BMI-PA random-intercept covariance by SES"
  ))

  sens_eq <- lavaan::lavaan(
    model = sub("riBMI ~~ riPA", "riBMI ~~ c(rc, rc, rc)*riPA", ri_clpm_syntax, fixed = TRUE),
    data = wide_sens, group = "ses", group.label = c("ses1", "ses2", "ses3"),
    int.ov.free = TRUE, meanstructure = TRUE,
    estimator = "MLR", se = "robust", missing = "fiml"
  )
  sens_lrt <- lavaan::lavTestLRT(sens_grouped, sens_eq,
    method = "satorra.2000",
    model.names = c("riBMI~~riPA free", "riBMI~~riPA equal")
  )
  sens_lrt_df <- cbind(model = rownames(sens_lrt), as.data.frame(sens_lrt))
  utils::write.csv(sens_lrt_df, "tables/bmi_sensitivity_moderation_lrt.csv", row.names = FALSE)
  print(knitr::kable(sens_lrt_df,
    row.names = FALSE, digits = 3,
    caption = "sensitivity: moderation test with flagged BMI set missing"
  ))
}
```

## 8 session information

```{r session-info}
sessionInfo()
```
